MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Model Selection

### 2020-04-17 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove commented code

# # Define biomart object
# mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
#                 host = "uswest.ensembl.org", # Ensembl Release 99 (January 2020)
#                 dataset = "hsapiens_gene_ensembl")
# 
# # Query biomart
# Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
#                        filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
#                        mart = mart)
Ensemble2HGNC = downloadFile("syn21907981")
GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::left_join(backgroundGenes) %>%
  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
```

<!-- ### Sex Check -->
<!-- ```{r Reported_Gender_plots} -->
<!-- counts$gene_id = rownames(counts)  -->
<!-- REPORTED.GENDER.COUNTS = GENE.GC.CONT %>%  -->
<!--   left_join(counts) %>% -->
<!--   dplyr::select(-one_of("percentage_gene_gc_content")) %>% -->
<!--   filter(chromosome_name == "X" |chromosome_name == "Y") %>%  -->
<!--   tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>% -->
<!--   dplyr::mutate(value = log(value)) %>% -->
<!--   dplyr::rename(`counts(log)`= value) %>%  -->
<!--   dplyr::rename(SampleID = item) %>% -->
<!--   left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>%  -->
<!--   dplyr::rename(`Reported Gender` = Reported_Gender)  -->

<!-- ##XIST and UTY expression  -->
<!-- #ENSG00000229807.11 and ENSG00000183878.15  -->
<!-- FILT <- REPORTED.GENDER.COUNTS %>%  -->
<!--   filter(gene_id == "ENSG00000229807.11" | gene_id == "ENSG00000183878.15") %>%  -->
<!--   dplyr::select(-one_of("chromosome_name")) %>%  -->
<!--   tidyr::spread(key = gene_id, value = `counts(log)`) %>%  -->
<!--   mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>%  -->
<!--   mutate(UTY = as.numeric(`ENSG00000183878.15`)) -->

<!-- p = ggplot(FILT, aes (x= XIST, y = UTY))  -->
<!-- p = p + geom_point(aes(color=`Reported Gender`))  -->
<!-- p -->
<!-- ``` -->
### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
ContCovariates <- c("ageOfDeath", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads","AlignmentRate", 
                     "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
<!-- Correlation/association between covariates at an FDR <= 0.1 -->
<!-- ```{r covariates.correlation, fig.width=10, fig.height=8} -->
<!-- COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05) -->
<!-- tmp = COVARIATES.CORRELATION$ESTIMATE -->
<!-- tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0 -->
<!-- h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate') -->
<!-- ComplexHeatmap::draw(h, heatmap_legend_side = 'left') -->
<!-- ``` -->

<!-- ## Explore metatdata -->
<!-- ```{r data.explore, fig.width = 12, fig.height = 12} -->
<!-- my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5)) -->

<!-- # RIN -->
<!-- p = list() -->
<!-- p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot() -->
<!-- p[[1]] = p[[1]] + ggtitle('RIN') + my.theme -->

<!-- # Age of Death -->
<!-- p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = ageOfDeath)) + geom_boxplot() -->
<!-- p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme -->

<!-- # PMI -->
<!-- p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot() -->
<!-- p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme -->

<!-- # Intronic Rate -->
<!-- p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot() -->
<!-- p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme -->

<!-- # IntergenicRate -->
<!-- p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot() -->
<!-- p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme -->

<!-- # Transcripts Detected -->
<!-- p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntragenicRate)) + geom_boxplot() -->
<!-- p[[6]] = p[[6]] + ggtitle('Intragenic Rate') + my.theme -->

<!-- # Mapped Reads -->
<!-- p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot() -->
<!-- p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme -->

<!-- # PercentAligned -->
<!-- p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TotalReads)) + geom_boxplot() -->
<!-- p[[8]] = p[[8]] + ggtitle('Total Reads') + my.theme -->

<!-- # rRNARate -->
<!-- p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot() -->
<!-- p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme -->

<!-- multiplot(plotlist = p, cols = 3) -->

<!-- # Institution -->
<!-- # vcd::mosaic(~ Institution + Dx.Tissue, data = COVARIATES) -->

<!-- # Gender -->
<!-- # vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES) -->
<!-- ``` -->
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
counts$gene_id = NULL

@@

<!-- ### Co-expression distribution -->
<!-- Coexpression of genes  -->
<!-- ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5} -->
<!-- cr = cor(t(CQN.GENE_EXPRESSION$E)) -->
<!-- hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation') -->
<!-- ``` -->

<!-- ### Sample clustering -->
<!-- ```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE} -->
<!-- # Find principal components of expression to plot -->
<!-- PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T) -->

<!-- # Plot first 2 PCs -->
<!-- plotdata <- data.frame(SampleID=rownames(PC$rotation),  -->
<!--                        PC1=PC$rotation[,1],  -->
<!--                        PC2=PC$rotation[,2]) -->

<!-- plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>% -->
<!--   tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') -->

<!-- p <- ggplot(plotdata, aes(x=PC1, y=PC2)) -->
<!-- p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath)) -->
<!-- p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.) -->
<!-- # p <- p + geom_text(aes(label= SampleID), size=4, hjust=0) -->
<!-- p -->
<!-- ``` -->
<!-- Tree based clustering of samples -->
<!-- ```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'} -->
<!-- # Eucledian tree based analysis -->
<!-- COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue")]) -->
<!-- COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0 -->

<!-- tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E))) -->
<!-- cols = WGCNA::labels2colors(COVARIATES.tmp); -->

<!-- WGCNA::plotDendroAndColors(tree,  -->
<!--                            colors = cols,  -->
<!--                            dendroLabels = FALSE,  -->
<!--                            abHeight = 0.80,  -->
<!--                            main = "Sample dendrogram", -->
<!--                            groupLabels = colnames(COVARIATES.tmp)) -->
<!-- ``` -->
<!-- ```{r temp, include = F} -->
<!-- dev.off() -->
<!-- gc() -->
<!-- ``` -->

<!-- ### Distribution of samples (log cpm) -->
<!-- ```{r lcpm.dist, cache=FALSE} -->
<!-- # Plot abberent distribution of logcpm counts -->
<!-- tmp1 = (CQN.GENE_EXPRESSION$E) %>% -->
<!--   rownameToFirstColumn('gene_id') %>% -->
<!--   tidyr::gather(SampleID, logCPM, -gene_id) %>% -->
<!--   left_join(COVARIATES %>% -->
<!--               rownameToFirstColumn('SampleID')) -->

<!-- p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density()  -->
<!-- p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx.Tissue, scale = 'free') -->
<!-- p -->
<!-- ``` -->

@@

<!-- #### Phase I (clinical, ancestry and sample specific technical variables) -->
<!-- ```{r phase1} -->
<!-- # Fit a base model with the following covariates -->
<!-- base_covars <- "Dx.Tissue" -->
<!-- random_effect <- "Individual_ID" -->

<!-- # Set base formula -->
<!-- base_formula <- glue::glue("~ ", glue::glue_collapse(base_covars, sep = " + ")) %>%  -->
<!--   glue::glue(" + (1|{random_effect})") -->

<!-- # Stepwise regression -->
<!-- phase1 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN, -->
<!--                          baseFormula = base_formula,  -->
<!--                          data = COVARIATES, -->
<!--                          variables = array(c("(1|Institution)", "(1|Reported_Gender)", "PMI",  -->
<!--                                                              "ageOfDeath", "RIN", "RIN2", "EV.1",  -->
<!--                                                              "EV.2", "EV.3", "EV.4", "EV.5"))) -->
<!-- phase1 -->
<!-- ``` -->

<!-- #### Phase II (batch effects) -->
<!-- ```{r phase2} -->
<!-- phase2 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN, -->
<!--                          baseFormula = phase1$formula,  -->
<!--                          data = COVARIATES, -->
<!--                          variables = array(c("(1|LibraryBatch)", "(1|FlowcellBatch)"))) -->
<!-- phase2 -->
<!-- ``` -->

@@

<!-- ### Store files in synapse -->
<!-- ```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE} -->
<!-- # Set annotations -->
<!-- all.annotations = list( -->
<!--   dataType = 'geneExpression', -->
<!--   dataSubtype = 'processed', -->
<!--   assay	 = 'rnaSeq', -->

<!--   tissue	= 'dorsolateral prefrontal cortex, anterior cingulate cortex',  -->
<!--   study = 'CMC',  -->

<!--   species = 'Human', -->
<!--   consortium	= 'CMC', -->

<!--   analysisType	= "data normalization", -->

<!--   transcriptNormalizationMethod	= 'CQN', -->
<!--   transcriptQuantificationMethod = 'featureCounts', -->
<!--   referenceSet = 'GRCh38' -->
<!-- ) -->

# Code
``` {r save_code}
CODE <- Folder(name = "test mvBIC - MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
CODE <- synStore(CODE)
```
<!-- # Store covariates -->
<!-- COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID') -->
<!-- write.table(COVARIATES, file = 'CMC_ACC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F) -->
<!-- COV_OBJ = synapser::File('CMC_ACC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id) -->
<!-- COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName,  -->
<!--                       executed = thisFile, activityDescription = activityDescription) -->
<!-- synSetAnnotations(COV_OBJ, annotations = all.annotations) -->

<!-- # Store filtered counts -->
<!-- PROCESSED_COUNTS$filteredExprMatrix$counts %>% -->
<!--   rownameToFirstColumn('ensembl_gene_id') %>% -->
<!--   write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F) -->
<!-- COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id) -->
<!-- COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName,  -->
<!--                       executed = thisFile, activityDescription = activityDescription) -->
<!-- synSetAnnotations(COUNT_OBJ, annotations = all.annotations) -->

<!-- # Store logCPM -->
<!-- CQN.GENE_EXPRESSION$E %>% -->
<!--   rownameToFirstColumn('ensembl_gene_id') %>% -->
<!--   write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F) -->
<!-- LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id) -->
<!-- LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName,  -->
<!--                       executed = thisFile, activityDescription = activityDescription) -->
<!-- synSetAnnotations(LCOUNT_OBJ, annotations = all.annotations) -->

<!-- # Store cqn offsets -->
<!-- CQN.GENE_EXPRESSION$offset %>% -->
<!--   rownameToFirstColumn('ensembl_gene_id') %>% -->
<!--   write.table(file = 'CMC_ACC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F) -->
<!-- OFFSET_OBJ = File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id) -->
<!-- OFFSET_OBJ = synStore(OFFSET_OBJ, used = ALL_USED_IDs, activityName = activityName,  -->
<!--                       executed = thisFile, activityDescription = activityDescription) -->
<!-- synSetAnnotations(OFFSET_OBJ, annotations = all.annotations) -->
<!-- ``` -->

<!-- |  *Results*                                  |  *SynapseID*                     | -->
<!-- |  -----------------------------------------  |   ---------                      | -->
<!-- |  Covariates                                 |  `r COV_OBJ$properties$id`       | -->
<!-- |  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     | -->
<!-- |  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    | -->
<!-- |  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    | -->

### Source Rmd
[Source markdown](`r thisFile`)

>>>>

Ensemble2HGNC = downloadFile("syn21907981")
GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::left_join(backgroundGenes) %>%
  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
```

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
ContCovariates <- c("ageOfDeath", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads","AlignmentRate", 
                     "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

@@

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```

### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
counts$gene_id = NULL

@@

@@

@@

# Code
``` {r save_code}
CODE <- Folder(name = "test mvBIC - MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
CODE <- synStore(CODE)
```
